---
title: "Normalization of Growth Metrics and Plan-Level Predictors"
author: "Donnie Minnick, Statistical Learning - Fall A 2025"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  word_document: default
  html_document:
    df_print: paged
subtitle: "A Modular Walkthrough of Scale Alignment, Transformation Logic, and Modeling Readiness"

---



This analysis leverages the following R packages: `dplyr`, `lubridate`, '`stringr` and `tidyr` for data manipulation, `knitr` for report formatting, and `ggplot2` and `ggridges` for visualization.

```{r echo = FALSE, message = FALSE}
library("dplyr")
library("ggplot2")
```

Load transformed Form 5500 data.

```{r}
plans <- readRDS("./data/plans_final.rds")
```

# Participant Growth Rate (PARTCP_GROWTH_RATE)

Flag outliers with the interquartile range method.

```{r}
PARTCP_GROWTH_RATE_check <- plans %>%
  mutate(q1 = quantile(PARTCP_GROWTH_RATE, 0.25, na.rm = TRUE),
         q3 = quantile(PARTCP_GROWTH_RATE, 0.75, na.rm = TRUE),
         iqr = q3 - q1,
         lower_bound = q1 - 1.5 * iqr,
         upper_bound = q3 + 1.5 * iqr,
         is_outlier = PARTCP_GROWTH_RATE < lower_bound | PARTCP_GROWTH_RATE > upper_bound)
```

Visualize outliers.

```{r}
ggplot(PARTCP_GROWTH_RATE_check, aes(x = PARTCP_GROWTH_RATE, fill = is_outlier)) +
  geom_histogram(binwidth = 0.05, position = "identity", alpha = 0.7) +
  scale_fill_manual(values = c("FALSE" = "steelblue", "TRUE" = "firebrick"),
                    labels = c("is_outlier" = "Is Outlier")) +
  labs(title = "Participant Growth Rate with Outlier Overlay",
       x = "Participant Growth Rate",
       y = "Plans") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        plot.title = element_text(size = 14, face = "bold")) 
```

The red bars on both tails immediately highlight plans with unusual participant growth rates. This visual makes it easy to narrate volatility across sectors or plan types.

The distribution is roughly bell-shaped and centered near zero, confirming that most plans are stable.



Diagnostic table.

```{r}
check_table <- check %>%
  group_by(SECTOR_TITLE) %>%
  summarise(
    outlier_count = sum(is_outlier, na.rm = TRUE),
    total = n(),
    pct_outliers = round(100 * outlier_count / total, 2)
  )

check_table
```

Structural clarity: See which sectors have tight vs. wide internal variation.

Outlier surfacing: Sectors with long tails or bimodal shapes may need fallback logic.

Fairness overlays: Helps narrate whether normalization preserves interpretability across sectors.

Sector-aware z score normalization

This centers each sector’s growth rate around its own mean.

Standard deviation is computed per sector, preserving internal variation.

norm_growth_sector is now scale-aligned but sector-relative.

```{r}
plans <- plans %>%
  group_by(SECTOR_TITLE) %>%
  mutate(
    PARTCP_GROWTH_RATE_sector_mean = mean(PARTCP_GROWTH_RATE, na.rm = TRUE),
    PARTCP_GROWTH_RATE_sector_sd   = sd(PARTCP_GROWTH_RATE, na.rm = TRUE),
    PARTCP_GROWTH_RATE_norm_growth_sector = (PARTCP_GROWTH_RATE - PARTCP_GROWTH_RATE_sector_mean) / PARTCP_GROWTH_RATE_sector_sd) %>%
  ungroup()
```

Sector aware distributions

```{r}
ggplot(plans, aes(x = PARTCP_GROWTH_RATE_norm_growth_sector)) +
  geom_density(fill = "steelblue", alpha = 0.6) +
  facet_wrap(~ SECTOR_TITLE, scales = "free_y") +
  labs(
    title = "Sector-Aware Normalized Growth Rate Distributions",
    x = "Normalized Growth Rate (Sector-Relative)",
    y = "Density"
  )


```

Centered within sector: Each sector’s mean is zero, so you're modeling relative growth rather than absolute.

Scaled within sector: Standard deviation is 1, which prevents dominant sectors from skewing regularization or distance-based models.

Preserves structural disparities: Sector-level volatility is retained, which is crucial for fairness-aware modeling.

Interpretable coefficientsA unit increase in this variable means “1 SD above sector mean,” which stakeholders can understand.
