---
title: "Normalization of Growth Metrics and Plan-Level Predictors"
author: "Donnie Minnick, Statistical Learning - Fall A 2025"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  word_document: default
  html_document:
    df_print: paged
subtitle: "A Modular Walkthrough of Scale Alignment, Transformation Logic, and Modeling Readiness"

---



This analysis leverages the following R packages: `dplyr`, `lubridate`, '`stringr` and `tidyr` for data manipulation, `knitr` for report formatting, and `ggplot2` and `ggridges` for visualization.

```{r echo = FALSE, message = FALSE}
library("dplyr")
library("ggplot2")
library("knitr")
```

Load transformed Form 5500 data.

```{r}
plans <- readRDS("../data/plans_final.rds")
```

# Participant Growth Rate (PARTCP_GROWTH_RATE)

Flag outliers with the interquartile range method.

```{r}
plans <- plans %>%
  mutate(
    PARTCP_GROWTH_RATE_q1 = quantile(PARTCP_GROWTH_RATE, 0.25, na.rm = TRUE),
    PARTCP_GROWTH_RATE_q3 = quantile(PARTCP_GROWTH_RATE, 0.75, na.rm = TRUE),
    PARTCP_GROWTH_RATE_iqr = PARTCP_GROWTH_RATE_q3 - PARTCP_GROWTH_RATE_q1,
    PARTCP_GROWTH_RATE_lower_bound = PARTCP_GROWTH_RATE_q1 - 1.5 * PARTCP_GROWTH_RATE_iqr,
    PARTCP_GROWTH_RATE_upper_bound = PARTCP_GROWTH_RATE_q3 + 1.5 * PARTCP_GROWTH_RATE_iqr,
    PARTCP_GROWTH_RATE_is_outlier = PARTCP_GROWTH_RATE < PARTCP_GROWTH_RATE_lower_bound |
                                    PARTCP_GROWTH_RATE > PARTCP_GROWTH_RATE_upper_bound)
```

Visualize outliers; zoom into dense region between -1 and 1.

```{r}
ggplot(plans, aes(x = PARTCP_GROWTH_RATE, fill = PARTCP_GROWTH_RATE_is_outlier)) +
  geom_histogram(binwidth = 0.05, position = "identity", alpha = 0.7) +
  scale_fill_manual(values = c("FALSE" = "steelblue", "TRUE" = "firebrick"),
                    labels = c("is_outlier" = "Is Outlier")) +
  coord_cartesian(xlim = c(-1, 1)) +
  labs(title = "Participant Growth Rate with Outlier Overlay",
       x = "Participant Growth Rate",
       y = "Plans") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        plot.title = element_text(size = 14, face = "bold")) 
```

Most retirement plans exhibit modest participant growth, forming a stable, bell-shaped distribution centered near zero. Outliers, highlighted in red, represent plans with unusually high or low growth rates, flagged for further review due to potential volatility, structural churn, or data anomalies.

Diagnostic table.

```{r}
check_table <- plans %>%
  group_by(SECTOR_TITLE) %>%
  summarise(outlier_count = sum(PARTCP_GROWTH_RATE_is_outlier, na.rm = TRUE),
            total = n(),
            pct_outliers = round(100 * outlier_count / total, 2))

kable(check_table,
      col.names = c("Sector", "Outlier Count", "Total", "% Outliers"),
      caption = "Outliers in Participant Growth Rate",
      format.args = list(big.mark = ","),
      align = c("l", "r", "r", "r"))
```

Outlier prevalence varies meaningfully across sectors, revealing structural patterns in participant growth volatility. Sectors such as Accommodation and Food Services, Administrative and Waste Management, and Agriculture exhibit the highest rates of outlier plans, likely reflecting episodic employment, seasonal churn, or fragmented plan structures. These elevated rates suggest the need for fallback-aware modeling and adjusted adequacy tier logic to avoid misclassification.

Mid-range volatility is observed in sectors like Information, Arts and Entertainment, and Professional Services, where outlier rates hover around 11–12%. These patterns may stem from startup dynamics, project-based employment, or creative churn, and warrant sector-aware normalization to preserve interpretability.

In contrast, sectors such as Education, Finance, Manufacturing, and Wholesale Trade show lower outlier rates (6–9%), indicating more stable participation patterns. These sectors may serve as anchors for benchmarking and fairness overlays.

Overall, the distribution of outliers across sectors underscores the importance of structural diagnostics in retirement plan analytics. Outlier-aware modeling and transparency in sector-level volatility are essential for ethical classification and stakeholder trust.

```{r}
plans <- plans %>%
  group_by(SECTOR_TITLE) %>%
  mutate(
    PARTCP_GROWTH_RATE_sector_mean = mean(PARTCP_GROWTH_RATE[!PARTCP_GROWTH_RATE_is_outlier], na.rm = TRUE),
    PARTCP_GROWTH_RATE_sector_sd   = sd(PARTCP_GROWTH_RATE[!PARTCP_GROWTH_RATE_is_outlier], na.rm = TRUE),
    PARTCP_GROWTH_RATE_norm_sector = if_else(!PARTCP_GROWTH_RATE_is_outlier,
      (PARTCP_GROWTH_RATE - PARTCP_GROWTH_RATE_sector_mean) / 
       PARTCP_GROWTH_RATE_sector_sd, NA_real_)) %>% ungroup()
```

Sector aware distributions.

```{r}
ggplot(plans, aes(x = PARTCP_GROWTH_RATE_norm_sector)) +
  geom_density(fill = "steelblue", alpha = 0.6) +
  facet_wrap(~ SECTOR_TITLE, scales = "free_y") +
  labs(title = "Sector-Aware Normalized Growth Rate Distributions",
       x = "Normalized Growth Rate (Sector-Relative)",
       y = "Density")
```

This grid of sector-specific density plots reveals how participant growth rates behave once normalized within each sector. By centering and scaling growth rates relative to sector norms, we uncover a consistent bell-shaped distribution across most industries — a sign that normalization has successfully neutralized scale disparities and volatility.

Sectors like Accommodation and Food Services and Transportation and Warehousing show wider spreads, hinting at structural churn or episodic employment patterns. In contrast, sectors such as Finance and Insurance or Educational Services display tighter distributions, suggesting more stable participation dynamics.

This visualization affirms that sector-aware normalization enables fair, interpretable comparisons across industries. It lays the groundwork for modeling, tiering, and fairness audits by ensuring that growth metrics reflect relative performance rather than raw magnitude.

# Average Contribution Per Participant (AVERAGE_CONTRIB_PER_PARTCP)

Flag outliers with the interquartile range method.

```{r}
plans <- plans %>%
  mutate(
    AVERAGE_CONTRIB_PER_PARTCP_q1 = quantile(AVERAGE_CONTRIB_PER_PARTCP, 0.25, na.rm = TRUE),
    AVERAGE_CONTRIB_PER_PARTCP_q3 = quantile(AVERAGE_CONTRIB_PER_PARTCP, 0.75, na.rm = TRUE),
    AVERAGE_CONTRIB_PER_PARTCP_iqr = AVERAGE_CONTRIB_PER_PARTCP_q3 - AVERAGE_CONTRIB_PER_PARTCP_q1,
    AVERAGE_CONTRIB_PER_PARTCP_lower_bound = AVERAGE_CONTRIB_PER_PARTCP_q1 - 1.5 * AVERAGE_CONTRIB_PER_PARTCP_iqr,
    AVERAGE_CONTRIB_PER_PARTCP_upper_bound = AVERAGE_CONTRIB_PER_PARTCP_q3 + 1.5 * AVERAGE_CONTRIB_PER_PARTCP_iqr,
    AVERAGE_CONTRIB_PER_PARTCP_is_outlier = AVERAGE_CONTRIB_PER_PARTCP < AVERAGE_CONTRIB_PER_PARTCP_lower_bound |
      AVERAGE_CONTRIB_PER_PARTCP > AVERAGE_CONTRIB_PER_PARTCP_upper_bound)
```

Visualize outliers; zoom into dense region between -1 and 1.

```{r}
ggplot(plans, aes(x = AVERAGE_CONTRIB_PER_PARTCP, fill = AVERAGE_CONTRIB_PER_PARTCP_is_outlier)) +
  geom_histogram(bins = 50, position = "identity", alpha = 0.7, color = "white") +
  scale_fill_manual(values = c("FALSE" = "steelblue", "TRUE" = "firebrick"),
                    labels = c("FALSE" = "Non-Outlier", "TRUE" = "Outlier"),
                    name = "Outlier Status") +
  scale_x_log10(breaks = c(1, 10, 100, 1000, 10000),
                labels = scales::comma_format()) +
  labs(title = "Average Contribution Per Participant with Outlier Overlay",
       x = "Average Contribution (Log Scale)",
       y = "Plan Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, hjust = 1),
        plot.title = element_text(size = 14, face = "bold"))
```

Most plans contribute between $100 and $1,000 per participant. Outliers (red) represent plans with unusually high contributions, which may reflect elite benefit structures or data anomalies.

Diagnostic table.

```{r}
diagnostic_table <- plans %>%
  group_by(SECTOR_TITLE) %>%
  summarize(outlier_count = sum(AVERAGE_CONTRIB_PER_PARTCP_is_outlier, na.rm = TRUE),
            total = n(),
            pct_outliers = round(100 * outlier_count / total, 1)) %>%
  arrange(desc(pct_outliers))

kable(diagnostic_table,
      col.names = c("Sector","Outlier Count", "Total", "% Outliers"),
      caption = "Outliers in Average Contribution Per Participant",
      format.args = list(big.mark = ","),
      align = c("l", "r", "r", "r"))
```

This table reveals sector-level disparities in participant contributions, meaning the amounts employees themselves choose to defer into their retirement plans. These aren’t employer matches or discretionary deposits — they reflect individual financial behavior, plan design, and structural capacity to save.

Sectors like Utilities (26.3% outliers), Professional Services (23.4%), and Educational Services (21.2%) show elevated rates of unusually high participant contributions. This suggests that employees in these industries may have greater financial literacy, higher disposable income, or access to plans that encourage aggressive deferral — such as 403(b)s or executive-tier 401(k)s.

High outlier prevalence in Information, Finance, and Management of Companies further supports the narrative of structurally privileged sectors, where employees may defer large portions of income due to compensation structure, plan flexibility, or cultural norms around retirement saving.

In contrast, sectors like Retail, Accommodation and Food Services, and Public Administration show lower outlier rates, indicating more uniform — and often modest — contribution behavior. These patterns may reflect wage constraints, limited plan features, or lower participation incentives.

```{r}
plans <- plans %>%
  group_by(SECTOR_TITLE) %>%
  mutate(AVERAGE_CONTRIB_PER_PARTCP_sector_mean = 
           mean(AVERAGE_CONTRIB_PER_PARTCP[!AVERAGE_CONTRIB_PER_PARTCP_is_outlier], na.rm = TRUE),
         AVERAGE_CONTRIB_PER_PARTCP_sector_sd = 
           sd(AVERAGE_CONTRIB_PER_PARTCP[!AVERAGE_CONTRIB_PER_PARTCP_is_outlier], na.rm = TRUE),
         AVERAGE_CONTRIB_PER_PARTCP_norm_sector = 
           if_else(!AVERAGE_CONTRIB_PER_PARTCP_is_outlier,
             (AVERAGE_CONTRIB_PER_PARTCP - AVERAGE_CONTRIB_PER_PARTCP_sector_mean) / 
              AVERAGE_CONTRIB_PER_PARTCP_sector_sd, NA_real_)) %>% ungroup()
```

Sector aware distributions.

```{r}
plans %>%
  filter(!is.na(AVERAGE_CONTRIB_PER_PARTCP_norm_sector)) %>%
ggplot(aes(x = AVERAGE_CONTRIB_PER_PARTCP_norm_sector)) +
  geom_density(fill = "steelblue", alpha = 0.6) +
  coord_cartesian(xlim = c(-3, 3)) +
  facet_wrap(~ SECTOR_TITLE, scales = "free_y") +
  labs(title = "Sector-Aware Normalized Participant Contribution Distributions",
       x = "Normalized Average Contribution (Sector-Relative)",
       y = "Density")
```

These plots reveal the distribution of normalized participant contributions across sectors, with each curve centered around the sector mean. Because these are employee deferrals, not employer matches, the shapes reflect individual saving behavior relative to sector norms.

Most sectors show left-skewed distributions, indicating that the majority of participants contribute below their sector’s average. This suggests that while high deferrals exist, they’re concentrated among a small subset — likely higher earners or financially literate participants. Sectors like Finance, Information, and Utilities show wider spreads, hinting at structural privilege or plan flexibility that enables aggressive saving.

In contrast, sectors like Retail, Accommodation, and Arts display tighter, more compressed distributions, suggesting limited deferral capacity or more uniform contribution behavior. These patterns may reflect wage constraints, plan design limitations, or cultural norms around retirement saving.

# Assets Per Participant (ASSETS_PER_PARTCP)

Flag outliers with the interquartile range method.

```{r}
plans <- plans %>%
  mutate(ASSETS_PER_PARTCP_q1 = quantile(ASSETS_PER_PARTCP, 0.25, na.rm = TRUE),
         ASSETS_PER_PARTCP_q3 = quantile(ASSETS_PER_PARTCP, 0.75, na.rm = TRUE),
         ASSETS_PER_PARTCP_iqr = ASSETS_PER_PARTCP_q3 - ASSETS_PER_PARTCP_q1,
         ASSETS_PER_PARTCP_lower_bound = ASSETS_PER_PARTCP_q1 - 1.5 * ASSETS_PER_PARTCP_iqr,
         ASSETS_PER_PARTCP_upper_bound = ASSETS_PER_PARTCP_q3 + 1.5 * ASSETS_PER_PARTCP_iqr,
         ASSETS_PER_PARTCP_is_outlier = ASSETS_PER_PARTCP < ASSETS_PER_PARTCP_lower_bound |
                                    ASSETS_PER_PARTCP > ASSETS_PER_PARTCP_upper_bound)
```

Visualize outliers; zoom into dense region between -1 and 1.

```{r}
ggplot(plans, aes(x = ASSETS_PER_PARTCP, fill = ASSETS_PER_PARTCP_is_outlier)) +
  geom_histogram(bins = 50, position = "identity", alpha = 0.7, color = "white") +
  scale_fill_manual(
    values = c("FALSE" = "steelblue", "TRUE" = "firebrick"),
    labels = c("FALSE" = "Non-Outlier", "TRUE" = "Outlier"),
    name = "Outlier Status"
  ) +
scale_x_log10(
  breaks = c(1, 10, 100, 1000, 10000, 100000, 1000000),
  labels = scales::comma_format()
) +
  labs(
    title = "Assets Per Participant with Outlier Overlay",
    x = "Assets (Log Scale)",
    y = "Plan Count"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 1),
    plot.title = element_text(size = 14, face = "bold")
  )

```

The distribution of assets per participant is heavily right-skewed, even on a log scale. That means most retirement plans cluster around modest asset levels per participant. A small subset of plans hold exceptionally high assets, flagged as outliers in red.

The log scale compresses extreme values, but the separation between typical and outlier plans remains visually stark. This pattern suggests structural stratification in retirement plan wealth.  Non-outlier plans likely reflect typical employer-sponsored 401(k)s with moderate balances, constrained by wage levels, tenure, and contribution behavior. 

Outlier plans may represent executive-tier plans, legacy pensions, or plans with concentrated ownership where a few participants hold disproportionately large balances.

```{r}
plans <- plans %>%
  group_by(SECTOR_TITLE) %>%
  mutate(ASSETS_PER_PARTCP_sector_mean = 
           mean(ASSETS_PER_PARTCP[!ASSETS_PER_PARTCP_is_outlier], na.rm = TRUE),
         ASSETS_PER_PARTCP_sector_sd = 
           sd(ASSETS_PER_PARTCP[!ASSETS_PER_PARTCP_is_outlier], na.rm = TRUE),
         ASSETS_PER_PARTCP_norm_sector = 
           if_else(!ASSETS_PER_PARTCP_is_outlier,
             (ASSETS_PER_PARTCP - ASSETS_PER_PARTCP_sector_mean) / 
              ASSETS_PER_PARTCP_sector_sd, NA_real_)) %>% ungroup()
```

Sector aware distributions.

```{r}
plans %>%
  filter(!is.na(ASSETS_PER_PARTCP_norm_sector)) %>%
ggplot(aes(x = ASSETS_PER_PARTCP_norm_sector)) +
  geom_density(fill = "steelblue", alpha = 0.6) +
  coord_cartesian(xlim = c(-3, 3)) +
  facet_wrap(~ SECTOR_TITLE, scales = "free_y") +
  labs(title = "Sector-Aware Normalized Assets Per Paticipant Distributions",
       x = "Normalized Average Contribution (Sector-Relative)",
       y = "Density")
```

Each density plot shows the distribution of assets per participant, normalized relative to the sector mean and standard deviation. This normalization allows for fair comparison across sectors with vastly different wage structures, tenure norms, and plan generosity.

Most sectors exhibit left-skewed distributions, even after normalization — meaning the majority of participants fall below their sector’s average asset level. This suggests that retirement wealth is concentrated among a minority, even within structurally privileged sectors.

Sectors like Finance, Utilities, and Management of Companies show wider spreads and heavier tails, indicating greater internal inequality in participant asset levels. These may reflect executive-heavy plans, concentrated ownership, or legacy balances.

In contrast, sectors like Retail, Accommodation, and Arts display tighter, compressed distributions, suggesting more uniform — and often modest — asset accumulation. These sectors may face structural constraints: lower wages, higher turnover, and limited plan engagement.

# Total Assets Growth Rate (TOTAL_ASSETS_GROWTH_RATE)

Flag outliers with the interquartile range method.

```{r}
plans <- plans %>%
  mutate(TOTAL_ASSETS_GROWTH_RATE_q1 = quantile(TOTAL_ASSETS_GROWTH_RATE, 0.25, na.rm = TRUE),
         TOTAL_ASSETS_GROWTH_RATE_q3 = quantile(TOTAL_ASSETS_GROWTH_RATE, 0.75, na.rm = TRUE),
         TOTAL_ASSETS_GROWTH_RATE_iqr = TOTAL_ASSETS_GROWTH_RATE_q3 - TOTAL_ASSETS_GROWTH_RATE_q1,
         TOTAL_ASSETS_GROWTH_RATE_lower_bound = TOTAL_ASSETS_GROWTH_RATE_q1 - 1.5 * TOTAL_ASSETS_GROWTH_RATE_iqr,
         TOTAL_ASSETS_GROWTH_RATE_upper_bound = TOTAL_ASSETS_GROWTH_RATE_q3 + 1.5 * TOTAL_ASSETS_GROWTH_RATE_iqr,
         TOTAL_ASSETS_GROWTH_RATE_is_outlier = TOTAL_ASSETS_GROWTH_RATE < TOTAL_ASSETS_GROWTH_RATE_lower_bound |
                                    TOTAL_ASSETS_GROWTH_RATE > TOTAL_ASSETS_GROWTH_RATE_upper_bound)
```

Visualize outliers; zoom into dense region between -1 and 1.

```{r}
ggplot(plans, aes(x = TOTAL_ASSETS_GROWTH_RATE, fill = TOTAL_ASSETS_GROWTH_RATE_is_outlier)) +
  geom_histogram(binwidth = 0.05, position = "identity", alpha = 0.7, color = "white") +
  scale_fill_manual(
    values = c("FALSE" = "steelblue", "TRUE" = "firebrick"),
    labels = c("FALSE" = "Non-Outlier", "TRUE" = "Outlier"),
    name = "Outlier Status"
  ) +
    coord_cartesian(xlim = c(-1, 2)) +
  labs(
    title = "Total Assets Growth Rate with Outlier Overlay",
    x = "Assets",
    y = "Plan Count"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 1),
    plot.title = element_text(size = 14, face = "bold")
  )

```

The distribution is roughly bell-shaped, centered around a growth rate of zero, suggesting that most plans experience modest asset changes year-over-year. This symmetry implies a mix of plans with slight gains and slight losses, consistent with normal market fluctuations and participant behavior.

The outliers, flagged in red, appear on both tails — representing plans with either exceptionally high growth or sharp declines. These could stem from:

High-growth outliers: New contributions, mergers, or asset transfers.

Negative-growth outliers: Heavy distributions, market losses, or plan terminations.

Outlier exclusion is essential for normalization and modeling — these plans behave fundamentally differently.

```{r}
plans <- plans %>%
  group_by(SECTOR_TITLE) %>%
  mutate(TOTAL_ASSETS_GROWTH_RATE_sector_mean = 
           mean(TOTAL_ASSETS_GROWTH_RATE[!TOTAL_ASSETS_GROWTH_RATE_is_outlier], na.rm = TRUE),
         TOTAL_ASSETS_GROWTH_RATE_sector_sd = 
           sd(TOTAL_ASSETS_GROWTH_RATE[!TOTAL_ASSETS_GROWTH_RATE_is_outlier], na.rm = TRUE),
         TOTAL_ASSETS_GROWTH_RATE_norm_sector = 
           if_else(!TOTAL_ASSETS_GROWTH_RATE_is_outlier,
             (TOTAL_ASSETS_GROWTH_RATE - TOTAL_ASSETS_GROWTH_RATE_sector_mean) / 
              TOTAL_ASSETS_GROWTH_RATE_sector_sd, NA_real_)) %>% ungroup()
```

Sector aware distributions.

```{r}
plans %>%
  filter(!is.na(TOTAL_ASSETS_GROWTH_RATE_norm_sector)) %>%
ggplot(aes(x = TOTAL_ASSETS_GROWTH_RATE_norm_sector)) +
  geom_density(fill = "steelblue", alpha = 0.6) +
  coord_cartesian(xlim = c(-3, 3)) +
  facet_wrap(~ SECTOR_TITLE, scales = "free_y") +
  labs(title = "Sector-Aware Normalized Total Assets Growth Rate Distributions",
       x = "Normalized Total Assets Growth Rate (Sector-Relative)",
       y = "Density")
```

Each density plot shows the distribution of normalized asset growth rates, centered around the sector mean. This normalization allows you to compare volatility and skew across sectors with vastly different baseline growth expectations.

Most sectors exhibit tight, symmetric distributions, suggesting that asset growth is relatively stable and centered around modest year-over-year change. This is especially true in sectors like Manufacturing, Retail, and Health Care, where plan behavior may be more regulated or predictable.

However, sectors like Finance, Information, and Management of Companies show wider spreads and heavier tails, indicating greater internal volatility. These may reflect plans with concentrated ownership, discretionary contributions, or episodic asset transfers.

Sectors like Arts, Accommodation, and Agriculture show compressed distributions, possibly due to smaller plan sizes, lower engagement, or structural constraints on growth.

# Loan Leakage Ratio (LOAN_LEAKAGE_RATIO)

Flag outliers with the interquartile range method.

```{r}
plans <- plans %>%
  mutate(LOAN_LEAKAGE_RATIO_q1 = quantile(LOAN_LEAKAGE_RATIO, 0.25, na.rm = TRUE),
         LOAN_LEAKAGE_RATIO_q3 = quantile(LOAN_LEAKAGE_RATIO, 0.75, na.rm = TRUE),
         LOAN_LEAKAGE_RATIO_iqr = LOAN_LEAKAGE_RATIO_q3 - LOAN_LEAKAGE_RATIO_q1,
         LOAN_LEAKAGE_RATIO_lower_bound = LOAN_LEAKAGE_RATIO_q1 - 1.5 * LOAN_LEAKAGE_RATIO_iqr,
         LOAN_LEAKAGE_RATIO_upper_bound = LOAN_LEAKAGE_RATIO_q3 + 1.5 * LOAN_LEAKAGE_RATIO_iqr,
         LOAN_LEAKAGE_RATIO_is_outlier = LOAN_LEAKAGE_RATIO < LOAN_LEAKAGE_RATIO_lower_bound |
                                    LOAN_LEAKAGE_RATIO > LOAN_LEAKAGE_RATIO_upper_bound)
```

Visualize outliers; zoom into dense region between -1 and 1.

```{r}
ggplot(plans, aes(x = LOAN_LEAKAGE_RATIO, fill = LOAN_LEAKAGE_RATIO_is_outlier)) +
  geom_histogram(binwidth = 0.01, position = "identity", alpha = 0.7, color = "white") +
  scale_fill_manual(
    values = c("FALSE" = "steelblue", "TRUE" = "firebrick"),
    labels = c("FALSE" = "Non-Outlier", "TRUE" = "Outlier"),
    name = "Outlier Status"
  ) +
    coord_cartesian(xlim = c(0, 0.3)) +
  labs(
    title = "Loan Leakage Ratio with Outlier Overlay",
    x = "Ratio",
    y = "Plan Count"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 1),
    plot.title = element_text(size = 14, face = "bold")
  )

```

Sharp concentration near zero; most plans have minimal leakage, which is reassuring from an adequacy and retention standpoint.

The red bars stretch into higher ratios, suggesting episodic or structural leakage, possibly due to aggressive loan policies, hardship withdrawals, or poor participant education.

Plans with high leakage ratios may be undermining participant adequacy even if contributions are strong.

Sector-level overlays could reveal whether certain industries (e.g., retail, hospitality) are structurally prone to leakage.

```{r}
plans <- plans %>%
  group_by(SECTOR_TITLE) %>%
  mutate(LOAN_LEAKAGE_RATIO_sector_mean = 
           mean(LOAN_LEAKAGE_RATIO[!LOAN_LEAKAGE_RATIO_is_outlier], na.rm = TRUE),
         LOAN_LEAKAGE_RATIO_sector_sd = 
           sd(LOAN_LEAKAGE_RATIO[!LOAN_LEAKAGE_RATIO_is_outlier], na.rm = TRUE),
         LOAN_LEAKAGE_RATIO_norm_sector = 
           if_else(!LOAN_LEAKAGE_RATIO_is_outlier,
             (LOAN_LEAKAGE_RATIO - LOAN_LEAKAGE_RATIO_sector_mean) / 
              LOAN_LEAKAGE_RATIO_sector_sd, NA_real_)) %>% ungroup()
```

Sector aware distributions.

```{r}
plans %>%
  filter(!is.na(LOAN_LEAKAGE_RATIO_norm_sector)) %>%
ggplot(aes(x = LOAN_LEAKAGE_RATIO_norm_sector)) +
  geom_density(fill = "steelblue", alpha = 0.6) +
  coord_cartesian(xlim = c(-3, 3)) +
  facet_wrap(~ SECTOR_TITLE, scales = "free_y") +
  labs(title = "Sector-Aware Normalized Loan Leakage Ratio Distributions",
       x = "Normalized Loan Leakage Ratio (Sector-Relative)",
       y = "Density")
```

Retail trade, accommodation, and food services show long right tails — suggesting episodic or structural leakage. These sectors may have more transient workforces or looser loan policies.

Finance, insurance, and public administration are tightly clustered — possibly due to stricter controls or higher participant literacy.

Utilities and mining show bimodal or skewed distributions — worth investigating whether plan design or participant behavior is driving that.
