---
title: "Classifying Retirement Plan Financial Adequacy Using Form 5500 Data"
author: "Donnie Minnick, Statistical Learning - Fall A 2025"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    df_print: paged
  word_document: default
subtitle: "Candidate Model Training and Comparison"
---

Logistic regression and random forest setup, cross-validation, fairness overlays, residual diagnostics, ROC/AUC curves, confusion matrices

# Load Libraries

This analysis leverages the following R packages: `dplyr`, `lubridate`, '`stringr` and `tidyr` for data manipulation, `knitr` for report formatting, and `ggplot2` and `ggridges` for visualization.

```{r echo = FALSE, message = FALSE}
library("dplyr")
library("ggplot2")
library("glmnet")
library("knitr")

source("../R/check_multicollinearity_factors.R")
source("../R/test_predictor_associations.R")
```

# Load Data

Load transformed Form 5500 data.

```{r}
plans <- readRDS("../data/plans_final.rds")
```

# Check Correlations

Check for correlations among predictors.

```{r check_correlations}
# Set vectors.

predictors <- c("PLAN_VINTAGE_GROUP",
                "SECTOR_TITLE_SHORT",
                "PARTCP_GROWTH_TIER",
                "CONTRIB_PARTCP_GROWTH_TIER",
                "CONTRIB_EMPLR_GROWTH_TIER",
                "LOAN_LEAKAGE_TIER",
                "TOTAL_ASSETS_GROWTH_TIER",
                "ASSETS_PER_PARTCP_TIER")

ordinal_vars <- c("PARTCP_GROWTH_TIER",
                  "CONTRIB_PARTCP_GROWTH_TIER",
                  "CONTRIB_EMPLR_GROWTH_TIER",
                  "LOAN_LEAKAGE_TIER",
                  "TOTAL_ASSETS_GROWTH_TIER",
                  "ASSETS_PER_PARTCP_TIER")

# ordinal_vars <- c("")

# Run the function to generate correlations summary.

assoc_summary <- test_predictor_associations(plans, predictors, ordinal_vars)

write.csv(assoc_summary, "assoc_summary.csv")

# rownames(assoc_summary) <- NULL
```

The Spearman correlation of 0.508 between CONTRIB_PARTCP_GROWTH_TIER and CONTRIB_EMPLR_GROWTH_TIER is almost certainly reflecting that matching dynamic: as employees contribute more, employers follow suit—up to the match cap. The association isn’t problematic—it’s structural, not spurious.

# Multicollinearity

Use the custom function `check_multicollinearity_factors` on the income dataset, using the specified predictors. Calculate Variance Inflation Factors (VIFs) to see how much each predictor is correlated with the others.

```{r check_multicollinearity}
predictors <- c("PLAN_VINTAGE_GROUP",
                "SECTOR_TITLE_SHORT",
                "PARTCP_GROWTH_TIER",
                "CONTRIB_PARTCP_GROWTH_TIER",
                "CONTRIB_EMPLR_GROWTH_TIER",
                "LOAN_LEAKAGE_TIER",
                "TOTAL_ASSETS_GROWTH_TIER",
                "ASSETS_PER_PARTCP_TIER")

vif_summary <- check_multicollinearity_factors(plans, predictors)

rownames(vif_summary) <- NULL

kable(vif_summary, 
      col.names = c("Variable", "VIF"),
      format.args = list(big.mark = ","),
      align = c("l", "r"))
```

Split data.

```{r}
set.seed(91210)

# Initial train/test split.

train_idx <- caret::createDataPartition(plans$ADEQUACY_IND, p = 0.7, list = FALSE)

plans_train <- plans[train_idx, ]

temp  <- plans[-train_idx, ]

# Split remaining into validation/test.

valid_idx <- caret::createDataPartition(temp$ADEQUACY_IND, p = 0.5, list = FALSE)

plans_validate <- temp[valid_idx, ]

plans_test <- temp[-valid_idx, ]
```

Check class balance in the train, validation and test sets.

```{r}
# Check balance function.

check_balance <- function(df, name) {
  df %>%
    count(ADEQUACY_IND) %>%
    mutate(prop = round(n / sum(n), 2),
           dataset = name) %>%
    select(dataset,
           ADEQUACY_IND, 
           n,
           prop)
}

# Generate data frame.

check <- bind_rows(check_balance(plans_train, "Train"),
                   check_balance(plans_validate, "Validation"),
                   check_balance(plans_test, "Test"))

# Produce table summary.

kable(check,
      col.names = c("Dataset", "Adequacy Level", "Count", "Percent"),
      caption = "Dataset Class Balance",
      format.args = list(big.mark = ","),
      align = c("l", "r", "r", "r"))

```




# Subset Selection with LASSO

```{r}
X <- model.matrix(ADEQUACY_IND ~ 
                    PLAN_VINTAGE_GROUP + 
                    SECTOR_TITLE_SHORT + 
                    PARTCP_GROWTH_TIER + 
                    CONTRIB_PARTCP_GROWTH_TIER + 
                    CONTRIB_EMPLR_GROWTH_TIER + 
                    LOAN_LEAKAGE_TIER + 
                    TOTAL_ASSETS_GROWTH_TIER + 
                    ASSETS_PER_PARTCP_TIER, 
                  data = plans)

y <- plans$ADEQUACY_IND

lasso_model <- cv.glmnet(X, y, alpha = 1)
```

```{r}

# Step 3: Extract coefficients at best lambda
coef_matrix <- coef(lasso_model, s = "lambda.min")

# Step 4: Convert to data frame and filter non-zero coefficients
selected_predictors <- as.data.frame(as.matrix(coef_matrix)) %>%
  rename(estimate = 1) %>%
  mutate(term = rownames(coef_matrix)) %>%
  filter(estimate != 0) %>%
  arrange(desc(abs(estimate)))

# Step 5: Preview selected predictors
selected_predictors


```

```{r}
coef_df <- as.data.frame(as.matrix(coef_matrix)) %>%
  rename(estimate = 1) %>%
  mutate(term = rownames(coef_matrix))

# Step 5: Filter eliminated predictors (estimate == 0)
eliminated_predictors <- coef_df %>%
  filter(estimate == 0)

# Step 6: Preview eliminated terms
print(eliminated_predictors$term)
```

# Subset Selection with Leaps

```{r}
library(leaps)

# Step 1: Fit subset selection model
subset_model <- regsubsets(ADEQUACY_IND ~ 
                             PLAN_VINTAGE_GROUP + 
                    SECTOR_TITLE_SHORT + 
                    PARTCP_GROWTH_TIER + 
                    CONTRIB_PARTCP_GROWTH_TIER + 
                    CONTRIB_EMPLR_GROWTH_TIER + 
                    LOAN_LEAKAGE_TIER + 
                    TOTAL_ASSETS_GROWTH_TIER + 
                    ASSETS_PER_PARTCP_TIER, , data = plans, nvmax = 20)

# Step 2: Extract model summary
subset_summary <- summary(subset_model)

# Step 3: Identify the best model size (lowest BIC)
best_size <- which.min(subset_summary$bic)

# Step 4: Extract logical vector of selected predictors
best_predictors_logical <- subset_summary$which[best_size, ]

# Step 5: Convert to character vector of predictor names
best_predictors <- names(best_predictors_logical[best_predictors_logical])  # includes intercept

# Step 6: Remove intercept if present
best_predictors <- setdiff(best_predictors, "(Intercept)")

# Step 7: Preview selected predictors
print(best_predictors)

```

# Logistic Model

```{r}
model_logistic <- glm(ADEQUACY_IND ~ 
                        PARTCP_GROWTH_TIER + 
                        CONTRIB_PARTCP_GROWTH_TIER + 
                        LOAN_LEAKAGE_TIER + 
                        TOTAL_ASSETS_GROWTH_TIER + 
                        ASSETS_PER_PARTCP_TIER,
                      data = plans, family = "binomial")

```


```{r}
summary(model_logistic)
```

Plot residuals.

```{r}
resid_df <- data.frame(fitted = fitted(model_logistic),
                       residuals = residuals(model_logistic, type = "pearson"),
                       actual = plans$ADEQUACY_IND)

# Convert actual outcome to factor for faceting.

resid_df$class <- factor(resid_df$actual, levels = c(0, 1), labels = c("Class 0", "Class 1"))

# Plot residuals vs. fitted by class.

ggplot(resid_df, aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.3, color = "blue") +
  geom_smooth(method = "loess", se = TRUE, color = "blue", linewidth = 0.8) +
  # facet_wrap(~ class) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted by Class",
       x = "Fitted Values",
       y = "Pearson Residuals") +
  theme_minimal()

```

Binned residuals

```{r}
arm::binnedplot(x = fitted(model_logistic),
         y = residuals(model_logistic, type = "response"),
         xlab = "Fitted Probabilities",
         ylab = "Average Residuals",
         main = "Binned Residual Plot")

```

Predict class scores.

```{r}
plans$predicted_prob <- predict(model_logistic, type = "response")
```

```{r}
threshold <- 0.5

plans$predicted_class <- ifelse(plans$predicted_prob > 0.5, 1, 0)

cm_logistic_train <- caret::confusionMatrix(factor(plans$predicted_class),
                                            factor(plans$ADEQUACY_IND))

cm_logistic_train

```

Generate ROC.

```{r}
roc_obj_log <- pROC::roc(plans$ADEQUACY_IND, plans$predicted_prob)
## Setting levels: control = 0, case = 1
## Setting direction: controls < cases
plot(roc_obj_log, col = "steelblue", main = "ROC Curve for Logistic Regression")

```

AUC.

```{r}
pROC::auc(roc_obj_log)
```

