---
title: "Classifying Retirement Plan Financial Adequacy Using Form 5500 Data"
author: "Donnie Minnick, Statistical Learning - Fall A 2025"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  md_document:
    variant: gfm
subtitle: "Model Training and Evaluation"
---

# Pipeline Scope 

This notebook encompasses and training and evaluation of logistic regression and random forest models.

# Load Libraries

This analysis leverages the following R packages: `dplyr`, `lubridate`, '`stringr` and `tidyr` for data manipulation, `knitr` for report formatting, and `ggplot2` and `ggridges` for visualization.

I use a customized R function to collect performance metrics across model runs.

```{r echo = FALSE, message = FALSE}
library("caret")
library("dplyr")
library("ggplot2")
library("glmnet")
library("knitr")
library("pROC")
library("randomForest")

source("../R/add_metrics_row.R")

# Initialize metrics data frame.

df_all_metrics <- NULL
```

# Load Data

Load training, validation and test sets.

```{r}
plans_train <- readRDS("../data/plans_train.rds")

plans_validate <- readRDS("../data/plans_validate.rds")

plans_test <- readRDS("../data/plans_test.rds")
```

# Train Models

## Logistic Regression Model

Produce logistic model and summary output.

```{r}
model_logistic <- glm(ADEQUACY_IND ~ 
                        SECTOR_TITLE_SHORT + 
                        PLAN_VINTAGE_GROUP + 
                        CONTRIB_EMPLR_GROWTH_TIER_Q + 
                        CONTRIB_PARTCP_GROWTH_TIER_Q + 
                        PARTCP_GROWTH_TIER_Q + 
                        ASSETS_PER_PARTCP_TIER_Q +
                        TOTAL_ASSETS_GROWTH_TIER_Q + 
                        LOAN_LEAKAGE_TIER_Q,
                      data = plans_train, family = "binomial")

summary(model_logistic)
```

Regarding coefficients with high p-values, this model was selected for predictive performance, not hypothesis testing. Some coefficients may not be statistically significant on their own, but they contribute to the model’s overall fit and help stabilize other effects. LASSO has already filtered out irrelevant predictors, so what remains reflects a balance of parsimony and predictive utility.

Plot residuals.

```{r}
resid_df <- data.frame(fitted = fitted(model_logistic),
                       residuals = residuals(model_logistic, type = "pearson"),
                       actual = model_logistic$model$ADEQUACY_IND)

# Convert actual outcome to factor for faceting.

resid_df$class <- factor(resid_df$actual, levels = c(0, 1), labels = c("Class 0", "Class 1"))

# Plot residuals vs. fitted by class.

ggplot(resid_df, aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.3, color = "blue") +
  geom_smooth(method = "loess", se = TRUE, color = "blue", linewidth = 0.8) +
  # facet_wrap(~ class) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted by Class",
       x = "Fitted Values",
       y = "Pearson Residuals") +
  theme_minimal()
```

The Pearson residual plot shows a mild curved pattern. This suggests some systematic deviation from model assumptions, possibly due to unmodeled nonlinearity or missing interactions. The residual variance is larger at the extremes (fitted values near 0 or 1), which is partly expected in logistic regression due to the changing variance structure of the binomial model. A few high residual points indicate potential outliers or influential observations.

Overall, the plot suggests the model captures the main relationship between predictors and the outcome, but may not fully account for all effects, leaving room for refinement.

Show binned residuals.

```{r}
arm::binnedplot(x = fitted(model_logistic),
         y = residuals(model_logistic, type = "response"),
         xlab = "Fitted Probabilities",
         ylab = "Average Residuals",
         main = "Binned Residual Plot")
```

The binned residual plot displays average residuals across bins of fitted probabilities, along with 95% simulation envelopes. Most binned residuals fall within the confidence bounds and fluctuate randomly around zero, with no consistent upward or downward trend. This indicates that the model’s predicted probabilities are well calibrated and that systematic bias is minimal. Minor deviations appear in the midrange probabilities (around 0.2–0.4), but they are small and within acceptable limits.

Overall, the plot supports that the logistic regression model fits the data adequately, with predictions that are unbiased on average.

Predict class scores on the training set.

```{r}
# Extract the model's training data.

model_data <- model_logistic$model

# Generate predicted probabilities.

model_data$predicted_prob <- predict(model_logistic, type = "response")

# Classify based on threshold.

model_data$predicted_class <- ifelse(model_data$predicted_prob > 0.5, 1, 0)

# Confusion matrix.

cm_logistic_train <- confusionMatrix(factor(model_data$predicted_class),
                                     factor(model_data$ADEQUACY_IND))

# Add logistic train metrics to metrics data frame.

df_all_metrics <- add_metrics_row(cm_logistic_train, "Logistic", "Train", df_all_metrics)

cm_logistic_train
```

The logistic regression model demonstrates strong, well-balanced classification performance. Accuracy (~89%) and kappa (~0.78) indicate robust predictive ability with minimal bias between classes. Sensitivity and specificity are both high and nearly equal, confirming the model distinguishes classes effectively. The non-significant McNemar’s test further supports that the model’s errors are symmetric. Overall, this is a well-calibrated, reliable classifier with room for only minor tuning.

Generate ROC.

```{r}
# Compute ROC curve
roc_obj_log <- pROC::roc(model_data$ADEQUACY_IND, model_data$predicted_prob)

# Plot ROC
plot(roc_obj_log, col = "steelblue", main = "ROC Curve for Logistic Regression")

```

The ROC curve shows strong model performance, with the curve rising well above the diagonal, indicating high sensitivity and specificity across thresholds. This suggests the model reliably distinguishes between adequate and inadequate plans.

Generate model AUC.

```{r}
pROC::auc(roc_obj_log)
```

An AUC of 0.9633 indicates excellent model performance. It means the logistic regression model can distinguish between adequate and inadequate plans with very high accuracy.

## Random Forest Model

Generate random forest model and show model performance.

```{r}
# plans_train$ADEQUACY_IND <- as.factor(plans_train$ADEQUACY_IND)

model_rf <- randomForest(ADEQUACY_IND ~ 
                           CONTRIB_EMPLR_GROWTH_TIER_Q + 
                           CONTRIB_PARTCP_GROWTH_TIER_Q +
                           PARTCP_GROWTH_TIER_Q + 
                           ASSETS_PER_PARTCP_TIER_Q + 
                           TOTAL_ASSETS_GROWTH_TIER_Q +
                           LOAN_LEAKAGE_TIER_Q,
                         data = plans_train,
                         ntree = 500,
                         mtry = 3,
                         importance = TRUE)

print(model_rf)
```

The random forest achieves strong and balanced classification accuracy (~84%) with no major class bias. It performs slightly below the logistic model’s 89% accuracy but benefits from flexibility and robustness to nonlinearity. Both models perform well; the choice depends on whether interpretability or pure predictive strength is the priority.

Generate a variable importance plot.

```{r}
varImpPlot(model_rf, type = 2, main = "Random Forest Variable Importance")
```

The random forest model confirms that contribution growth, especially participant-driven growth, is the strongest predictor of plan adequacy, followed closely by employer contributions and loan leakage. This resonates as industry studies confirm that those who contribute consistently to retirement savings - and do not draw early on savings - are generally more financially prepared for this life event.   

Asset-based tiers play a supporting role, while total asset growth shows the least influence.

Generate prediction on training data and confusion matrix.

```{r}
# Filter training data to match model input.

vars_used <- all.vars(formula(model_rf))

plans_train_rf <- na.omit(plans_train[, vars_used])

# Generate predictions.

rf_preds <- predict(model_rf, newdata = plans_train_rf, type = "response")

# Confusion matrix.

cm_rf_train <- confusionMatrix(rf_preds, plans_train_rf$ADEQUACY_IND, positive = "0")

df_all_metrics <- add_metrics_row(cm_rf_train, "RandomForest", "Train", df_all_metrics)

cm_rf_train
```

The random forest’s 96.7% test accuracy is unusually high relative to its 84% OOB accuracy, suggesting possible data leakage, overlap, or overly optimistic test sampling. I'll verify performance with validation and test sets.

Generate ROC.

```{r}
# Generate predicted probabilities for Class 0.

rf_probs <- predict(model_rf, newdata = plans_train_rf, type = "prob")[, "0"]

# Compute ROC using actuals and probabilities.

roc_obj <- pROC::roc(plans_train_rf$ADEQUACY_IND, rf_probs, levels = c("1", "0"), direction = "<")

# Plot ROC curve.

plot(roc_obj, col = "steelblue", main = "Random Forest ROC Curve")
```

When combined with the 96.7% test accuracy and 84% OOB accuracy, this near-perfect ROC curve reinforces that the model may be over-fitting the training data.

Generate model AUC.

```{r}
auc(roc_obj)
```

This AUC is signaling near-perfect discrimination, better than nearly all real-world models built on behavioral or operational data.  Need to check with validation data.

## Performance Metric Comparison

Show performance metric comparison for both models with training data.

```{r}
kable(df_all_metrics,
      col.names = c("Metric", names(df_all_metrics)[-1]),
      caption = "Model Performance Comparison (Wide Format)",
      format.args = list(big.mark = ","),
      align = c("l", rep("r", ncol(df_all_metrics) - 1)))
```

Random Forest may be overfitting on the training data, given its near-perfect scores. Logistic regression, while less precise, may offer better generalization and interpretability, especially if validation and test metrics show random forest performance dropping.

# Validate Models

Validate models with validation data.

## Logistic Regression Model

Predict outcomes with the validation data.

```{r}
pred_probs_logistic <- predict(model_logistic, newdata = plans_validate, type = "response")

class_pred_logistic <- ifelse(pred_probs_logistic >= 0.5, 1, 0)

cm_logistic_validate <- caret::confusionMatrix(factor(class_pred_logistic),
                                            factor(plans_validate$ADEQUACY_IND))

# Capture performance metrics.

df_all_metrics <- add_metrics_row(cm_logistic_validate, "Logistic", "Validate", df_all_metrics)

cm_logistic_validate
```

Based on these metrics, the logistic model generalizes well. The slight drop in specificity on the validation set (compared to the metrics for the training set), paired with a significant McNemar result, hints at a modest skew toward false positives, i.e. adequate plans mis-classified as inadequate.  The model is accurate overall, but the error distribution isn’t balanced. This contrasts with the training set McNemar result (0.5791), which showed no directional bias.

## Random Forest Model

```{r}
rf_preds <- predict(model_rf, newdata = plans_validate, type = "response")

# table(Predicted = rf_preds, Actual = plans_validate$ADEQUACY_IND)

cm_rf_validate <- confusionMatrix(rf_preds, plans_validate$ADEQUACY_IND, positive = "0")

# Capture performance metrics.

df_all_metrics <- add_metrics_row(cm_rf_validate, "RandomForest", "Validate", df_all_metrics)

cm_rf_validate
```

## Performance Metric Comparison

Show performance metric comparison for models with training and validation runs.

```{r}
kable(df_all_metrics,
      col.names = c("Metric", names(df_all_metrics)[-1]),
      caption = "Model Performance Comparison (Wide Format)",
      format.args = list(big.mark = ","),
      align = c("l", rep("r", ncol(df_all_metrics) - 1)))
```

# Test Models

## Logistic Regression Model

```{r}
pred_probs_logistic <- predict(model_logistic, newdata = plans_test, type = "response")

class_pred_logistic <- ifelse(pred_probs_logistic >= 0.5, 1, 0)

cm_logistic_test <- caret::confusionMatrix(factor(class_pred_logistic),
                                            factor(plans_test$ADEQUACY_IND))

# Capture performance metrics.

df_all_metrics <- add_metrics_row(cm_logistic_test, "Logistic", "Test", df_all_metrics)

cm_logistic_test
```

## Random Forest Model

```{r}
rf_preds <- predict(model_rf, newdata = plans_test, type = "response")

# table(Predicted = rf_preds, Actual = plans_validate$ADEQUACY_IND)

cm_rf_test <- confusionMatrix(rf_preds, plans_test$ADEQUACY_IND, positive = "0")

# Capture performance metrics.

df_all_metrics <- add_metrics_row(cm_rf_test, "RandomForest", "Test", df_all_metrics)

cm_rf_test
```

# Final Comparison

## Performance Metric Comparison

Show performance metric comparison for models with training and validation runs.

```{r}
kable(df_all_metrics,
      col.names = c("Metric", names(df_all_metrics)[-1]),
      caption = "Model Performance Comparison (Wide Format)",
      format.args = list(big.mark = ","),
      align = c("l", rep("r", ncol(df_all_metrics) - 1)))

saveRDS(df_all_metrics, "../data/model_metrics.rds")
```

# Misclassification Analysis

To assess whether the models were simply reproducing the engineered adequacy score, I conducted a misclassification analysis across adequacy tiers using the unseen test data. Both the logistic regression and random forest models showed clear signs of structural learning.

Tag misclassified plans with flags for false positives and false negatives using the test set.

```{r}
plans_test <- plans_test %>%
  mutate(ADEQUACY_IND_NUM = as.numeric(as.character(ADEQUACY_IND)),
         LOGISTIC_PRED_NUM = as.numeric(as.character(class_pred_logistic)),
         RF_PRED_NUM = as.numeric(as.character(rf_preds))) %>%
  mutate(LOGISTIC_FP = LOGISTIC_PRED_NUM == 1 & ADEQUACY_IND_NUM == 0,
         LOGISTIC_FN = LOGISTIC_PRED_NUM == 0 & ADEQUACY_IND_NUM == 1,
         RF_FP = RF_PRED_NUM == 1 & ADEQUACY_IND_NUM == 0,
         RF_FN = RF_PRED_NUM == 0 & ADEQUACY_IND_NUM == 1)
```

Summarize misclassifications by adequacy score to show whether misclassified plans are near the threshold (e.g., score of 3 or 4).

```{r}
class_check <- plans_test %>%
  group_by(ADEQUACY_SCORE) %>%
  summarise(
    Logistic_FN = sum(LOGISTIC_FN),
    Logistic_FP = sum(LOGISTIC_FP),
    RF_FN = sum(RF_FN),
    RF_FP = sum(RF_FP),
    Total = n()
  ) %>%
  arrange(ADEQUACY_SCORE)

kable(class_check,
      col.names = c("Adequacy Score", "Logistic_FN", "Logistic_FP", "RF_FN", "RF_FP", "Total"),
      caption = "Misclassifications by Model",
      format.args = list(big.mark = ","),
      align = c("l", "r", "r", "r", "r", "r"))
```

Misclassifications clustered around borderline scores, particularly at score 3 (just below the adequacy threshold) and score 4 (just above it). 

Both models flagged several score 3 plans as adequate and several score 4 plans as inadequate, suggesting they were responding to signals in tiered features rather than rigidly following the score cutoff. 

At the extremes (scores 0, 1, and 6), both models classified plans cleanly, reinforcing that they respected strong adequacy signals. 

These patterns validate that the models were not memorizing thresholds but learning nuanced structural patterns embedded in the feature space.

Create faceted bar plot.

```{r}
misclass_long <- class_check %>%
  tidyr::pivot_longer(cols = c(Logistic_FN, Logistic_FP, RF_FN, RF_FP),
               names_to = "Model_Error",
               values_to = "Count") %>%
  mutate(Model = case_when(grepl("^Logistic", Model_Error) ~ "Logistic",
                           grepl("^RF", Model_Error) ~ "Random Forest"),
         Error_Type = case_when(grepl("FN$", Model_Error) ~ "False Negative",
                                grepl("FP$", Model_Error) ~ "False Positive"))
ggplot(misclass_long, aes(x = factor(ADEQUACY_SCORE), y = Count, fill = Error_Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  # geom_vline(xintercept = 4.5, linetype = "dashed", color = "black", linewidth = 0.8) +
  facet_wrap(~ Model) +
  scale_fill_manual(values = c("steelblue", "lightsteelblue")) +
  labs(title = "Misclassifications by Adequacy Score",
       x = "Adequacy Score",
       y = "Count",
       fill = "Error Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
        plot.title = element_text(size = 14, face = "bold"))
```

# High-Level Takeaways

Summary of observations from performance metrics.

| Dimension | Logistic Regression | Random Forest | 
| --------- | ------------------- | ------------- |
| Train     |	Solid performance, slightly lower than RF	| Near-perfect fit, possible overfitting |
| Validate  | Strong generalization | Noticeable drop from train, less stable |
| Test      |	Consistent and balanced | Slight recovery, but still below logistic |

# Diagnostic Highlights

**Logistic Regression**

* Consistent performance metrics across splits; accuracy of ~89–90% (train/validate) and ~86% (test).
* Balanced accuracy and Kappa remain strong across sets, suggesting stable class separation and agreement.
* Sensitivity vs. specificity are well-matched, indicating no major bias toward either class.
* McNemar’s p-values near 1 for validation set suggest symmetric mis-classification, no systematic error.

**Random Forest**

* Training accuracy was high at 96.7%, suggesting possible over-fitting.
* Validation accuracy dropped to 82.8% and Kappa fell to 0.65; this suggests reduced generalization.
* Sensitivity greater than specificity across splits; RF may be slightly favoring the positive class (Class 0).
* McNemar’s P-Values are lower (0.76–0.63), but still not significant; mild asymmetry in errors.

# Conclusion

Logistic regression offers consistent, balanced performance across training, validation, and test sets, with strong agreement and minimal bias. Random forest shows excellent fit on training data but a notable drop in validation accuracy and agreement, suggesting potential over-fitting. While both models perform well, logistic regression offers greater stability and interpretability for deployment.
