---
title: "Classifying Retirement Plan Financial Adequacy Using Form 5500 Data"
author: "Donnie Minnick, Statistical Learning - Fall A 2025"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  md_document:
    variant: gfm
subtitle: "Model Training and Evaluation"
---

# Pipeline Scope 

This notebook encompasses and training and evaluation of logistic regression and random forest models.

# Load Libraries

This analysis leverages the following R packages: `dplyr`, `lubridate`, `stringr` and `tidyr` for data manipulation, `knitr` for report formatting, and `ggplot2` and `ggridges` for visualization.

I use a customized R functions to collect performance metrics across model runs and generate a common set of metrics for model evaluation.

```{r message = FALSE}
library("caret")
library("dplyr")
library("ggplot2")
library("glmnet")
library("knitr")
library("pROC")
library("randomForest")

source("../R/add_metrics_row.R")
source("../R/evaluate_model.R")

# Initialize metrics data frame.

df_all_metrics <- data.frame(Metric = character(), stringsAsFactors = FALSE)
```

# Load Data

Load training, validation and test sets.

```{r}
plans_train <- readRDS("../data/plans_train.rds")

plans_validate <- readRDS("../data/plans_validate.rds")

plans_test <- readRDS("../data/plans_test.rds")
```

# Train Models

## Logistic Regression Model

Produce logistic model and summary output.

```{r}
model_logistic <- glm(ADEQUACY_IND ~ 
                        SECTOR_TITLE_SHORT + 
                        PLAN_VINTAGE_GROUP + 
                        CONTRIB_EMPLR_GROWTH_TIER_Q + 
                        CONTRIB_PARTCP_GROWTH_TIER_Q + 
                        PARTCP_GROWTH_TIER_Q + 
                        ASSETS_PER_PARTCP_TIER_Q +
                        TOTAL_ASSETS_GROWTH_TIER_Q + 
                        LOAN_LEAKAGE_TIER_Q,
                      data = plans_train, family = "binomial")

summary(model_logistic)
```

This model was selected for predictive performance rather than inferential testing. While some coefficients exhibit high p-values, their inclusion reflects retained predictive utility post-LASSO regularization. These terms may contribute through interaction effects, variance stabilization, or sector-specific nuance not captured by significance alone. The final specification balances parsimony, interpretability, and generalization across adequacy tiers and sector stratifications.

Plot residuals.

```{r}
resid_df <- data.frame(fitted = fitted(model_logistic),
                       residuals = residuals(model_logistic, type = "pearson"),
                       actual = model_logistic$model$ADEQUACY_IND)

# Convert actual outcome to factor for faceting.

resid_df$class <- factor(resid_df$actual, levels = c(0, 1), labels = c("Class 0", "Class 1"))

# Plot residuals vs. fitted by class.

ggplot(resid_df, aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.3, color = "blue") +
  geom_smooth(method = "loess", se = TRUE, color = "blue", linewidth = 0.8) +
  # facet_wrap(~ class) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted by Class",
       x = "Fitted Values",
       y = "Pearson Residuals") +
  theme_minimal()
```

The Pearson residual plot reveals a mild curvature, suggesting potential nonlinearity or missing interactions not fully captured by the current specification. Residual variance increases near fitted values of 0 and 1, which aligns with the binomial variance structure in logistic regression. A few high residuals may indicate outliers or influential observations worth further review. Overall, the model captures the primary signal between predictors and adequacy, but the residual pattern suggests room for refinement—particularly in modeling nonlinear effects or sector-specific interactions.

Show binned residuals.

```{r}
arm::binnedplot(x = fitted(model_logistic),
         y = residuals(model_logistic, type = "response"),
         xlab = "Fitted Probabilities",
         ylab = "Average Residuals",
         main = "Binned Residual Plot")
```

Most binned residuals fall within the confidence bounds and fluctuate randomly around zero, with no consistent upward or downward trend. This indicates that the model’s predicted probabilities are well calibrated and that systematic bias is minimal. Some deviations appear in the midrange probabilities (around 0.2–0.4), but they are small and within acceptable limits.

Overall, the plot supports that the logistic regression model fits the data adequately, with predictions that are unbiased on average.

Predict class scores on the training set and get prediction results using the `evaluate_model` function.

```{r}
train_results_logistic <- evaluate_model(model_logistic, plans_train, "ADEQUACY_IND", "Logistic", "Train", df_all_metrics)

df_all_metrics <- train_results_logistic$metrics_df
```

Show results of the confusion matrix.

```{r}
train_results_logistic$metrics_df
```

The confusion matrix reveals strong model performance, with high accuracy, balanced sensitivity and specificity, and substantial agreement beyond chance (Kappa = 0.82). The Mcnemar test shows no significant directional bias, and detection rates align closely with prevalence. These diagnostics support the model’s reliability and fairness across adequacy tiers.

Show the ROC curve.

```{r}
plot(train_results_logistic$roc_obj,
     col = "steelblue",         
     lwd = 2,                   
     main = "ROC Curve - Logistic Model (Train Set)",
     print.auc = TRUE,          
     print.auc.cex = 1.2,       
     print.auc.col = "darkred",
     legacy.axes = TRUE)       
```

The ROC curve shows strong model performance, with the curve rising well above the diagonal, indicating high sensitivity and specificity across thresholds. This suggests the model reliably distinguishes between adequate and inadequate plans.

An AUC of 0.9701 indicates excellent model performance. It means the logistic regression model can distinguish between adequate and inadequate plans with very high accuracy.

## Random Forest Model

Generate random forest model and show model performance.

```{r}
# plans_train$ADEQUACY_IND <- as.factor(plans_train$ADEQUACY_IND)

model_rf <- randomForest(ADEQUACY_IND ~ 
                           CONTRIB_EMPLR_GROWTH_TIER_Q + 
                           CONTRIB_PARTCP_GROWTH_TIER_Q +
                           PARTCP_GROWTH_TIER_Q + 
                           ASSETS_PER_PARTCP_TIER_Q + 
                           TOTAL_ASSETS_GROWTH_TIER_Q +
                           LOAN_LEAKAGE_TIER_Q,
                         data = plans_train,
                         ntree = 500,
                         mtry = 3,
                         importance = TRUE)

print(model_rf)
```

This random forest model achieves ~86% accuracy with balanced error rates across adequacy classes. The OOB estimate supports generalization, and the confusion matrix shows no major classification bias.

Generate a variable importance plot.

```{r}
varImpPlot(model_rf, type = 2, main = "Random Forest Variable Importance")
```

The random forest model confirms that contribution growth, especially participant-driven growth, is the strongest predictor of plan adequacy, followed closely by employer contributions and loan leakage. This resonates as industry studies confirm that those who contribute consistently to retirement savings - and do not draw early on savings - are generally more financially prepared for this life event.   

Asset-based tiers play a supporting role, while total asset growth shows the least influence.

Predict class scores on the training set and get prediction results using the `evaluate_model` function.

```{r}
train_results_rf <- evaluate_model(model_rf, plans_train, "ADEQUACY_IND", "RF", "Train", df_all_metrics)

df_all_metrics <- train_results_rf$metrics_df
```

Show results of the confusion matrix.

```{r}
train_results_rf$metrics_df
```

This random forest model demonstrates exceptional performance, with high accuracy, balanced sensitivity and specificity, and strong agreement beyond chance (Kappa = 0.93). The Mcnemar test confirms no directional bias, and balanced accuracy supports fairness-aware classification across adequacy tiers. These diagnostics validate the model’s reliability and stakeholder readiness.

Show the ROC curve.

```{r}
plot(train_results_rf$roc_obj,
     col = "steelblue",         
     lwd = 2,                   
     main = "ROC Curve - Random Forest Model (Train Set)",
     print.auc = TRUE,          
     print.auc.cex = 1.2,       
     print.auc.col = "darkred",
     legacy.axes = TRUE)       
```

The near-perfect ROC curve, paired with 96% test accuracy and 84% OOB accuracy, suggests the model may be overfitting to training data.

The AUC reflects exceptional discrimination, outperforming most real-world models built on behavioral or operational data. While promising, this level of performance warrants validation on holdout data to confirm generalization and fairness across adequacy tiers.

# Validate Models

Evaluate the models with validation data.

## Logistic Regression Model

Predict class scores on the validation set and get prediction results using the `evaluate_model` function.

```{r}
validate_results_logistic <- evaluate_model(model_logistic, plans_validate, "ADEQUACY_IND", "Logistic", "Validate", df_all_metrics)

df_all_metrics <- validate_results_logistic$metrics_df
```

Show results of the confusion matrix.

```{r}
validate_results_logistic$metrics_df
```

The logistic model performs well on the validation set, with an AUC of 0.951 and balanced accuracy of 0.867, indicating strong class separation and fair treatment of positives and negatives. Sensitivity (0.886) slightly exceeds specificity (0.849), favoring detection of adequacy. Kappa (0.735) shows substantial agreement, and the McNemar p-value (0.728) confirms no directional bias in errors. Overall, the model generalizes well with no major fairness concerns.

Show the ROC curve.

```{r}
plot(validate_results_logistic$roc_obj,
     col = "steelblue",         
     lwd = 2,                   
     main = "ROC Curve - Logistic Model (Validate Set)",
     print.auc = TRUE,          
     print.auc.cex = 1.2,       
     print.auc.col = "darkred",
     legacy.axes = TRUE)       
```

The ROC curve for the logistic model on the validation set shows strong separation between classes, with an AUC of 0.951, indicating excellent discriminatory power. This suggests the model is highly effective.

## Random Forest Model

Predict class scores on the validation set and get prediction results.

```{r}
validate_results_rf <- evaluate_model(model_rf, plans_validate, "ADEQUACY_IND", "RF", "Validate", df_all_metrics)

df_all_metrics <- validate_results_rf$metrics_df
```

Show results of the confusion matrix.

```{r}
validate_results_rf$metrics_df
```

The random forest model shows decent validation performance with an AUC of 0.914 and balanced accuracy of 0.832, but it trails the logistic model across most metrics. The drop from training suggests overfitting, and while sensitivity and specificity are balanced, overall classification power is weaker. The McNemar p-value of 1.000 indicates no directional bias, but the model generalizes less effectively than logistic.

Show the ROC curve.

```{r}
plot(validate_results_rf$roc_obj,
     col = "steelblue",         
     lwd = 2,                   
     main = "ROC Curve - Random Forest Model (Validate Set)",
     print.auc = TRUE,          
     print.auc.cex = 1.2,       
     print.auc.col = "darkred",
     legacy.axes = TRUE)       
```

The ROC curve for the random forest model on the validation set shows solid classification performance, with the curve clearly above the diagonal, indicating better-than-random predictions. The AUC of 0.914 confirms strong discriminatory power, though it’s notably lower than the training AUC of 0.996 and slightly behind the logistic model’s validation AUC of 0.951. This drop suggests the random forest model may be overfitting to training data and generalizing less effectively. Still, the curve shape reflects balanced sensitivity and specificity across thresholds, with no major fairness concerns.

# Test Models

## Logistic Regression Model

Predict class scores on the test set and get prediction results.

```{r}
test_results_logistic <- evaluate_model(model_logistic, plans_test, "ADEQUACY_IND", "Logistic", "Test", df_all_metrics)

df_all_metrics <- test_results_logistic$metrics_df
```

Show results of the confusion matrix.

```{r}
test_results_logistic$metrics_df
```

The logistic model performs consistently well on the test set, with an AUC of 0.935 and balanced accuracy of 0.878, confirming strong generalization. Sensitivity (0.901) and specificity (0.856) remain well-balanced, and the Kappa score of 0.758 reflects substantial agreement. The McNemar p-value (0.584) shows no directional bias in errors. Overall, the model maintains high classification power and fairness across splits, validating its reliability for stakeholder-facing diagnostics.

Show the ROC curve.

```{r}
plot(test_results_logistic$roc_obj,
     col = "steelblue",         
     lwd = 2,                   
     main = "ROC Curve - Logistic Model (Test Set)",
     print.auc = TRUE,          
     print.auc.cex = 1.2,       
     print.auc.col = "darkred",
     legacy.axes = TRUE)       
```

The ROC curve for the logistic model on the test set shows strong classification performance, with the curve well above the diagonal, indicating reliable separation between adequacy classes. The AUC of 0.935 confirms excellent discriminatory power, validating the model’s ability to generalize beyond training and validation data. The curve’s shape reflects balanced sensitivity and specificity across thresholds, with no signs of directional bias or fairness concerns.

## Random Forest Model

Predict class scores on the test set and get prediction results.

```{r}
test_results_rf <- evaluate_model(model_rf, plans_test, "ADEQUACY_IND", "RF", "Test", df_all_metrics)

df_all_metrics <- test_results_rf$metrics_df
```

Show results of the confusion matrix.

```{r}
test_results_rf$metrics_df
```

The Random Forest model performs well on the test set, with an AUC of 0.923 and balanced accuracy of 0.855, confirming strong generalization. Sensitivity (0.863) and specificity (0.848) are well-balanced, and the Kappa score of 0.710 reflects substantial agreement. While slightly behind the logistic model in most metrics, the RF model maintains reliable classification power and fairness, with a McNemar p-value of 1.000 indicating no directional bias in errors.

Show the ROC curve.

```{r}
plot(test_results_rf$roc_obj,
     col = "steelblue",         
     lwd = 2,                   
     main = "ROC Curve - Random Forest Model (Test Set)",
     print.auc = TRUE,          
     print.auc.cex = 1.2,       
     print.auc.col = "darkred",
     legacy.axes = TRUE)       
```

The ROC curve for the random forest model on the test set shows strong classification performance, with the curve clearly above the diagonal, indicating reliable separation between adequacy classes. The AUC of 0.923 confirms high discriminatory power, though slightly behind its training AUC of 0.996 and the logistic model’s test AUC of 0.935. The curve reflects balanced sensitivity and specificity across thresholds, with no signs of directional bias. Overall, the RF model generalizes well and remains a solid performer.

# Final Comparison

The logistic model demonstrates consistent performance across train, validate, and test splits, with high AUC, balanced sensitivity/specificity, and strong interpretability, making it ideal for diagnostics and stakeholder transparency.

The random forest model performs well but shows signs of overfitting, with a sharper drop in validation metrics and slightly lower test performance. It offers robust classification but less interpretability, which may limit stakeholder trust in high-stakes contexts.

For applications prioritizing transparency and stakeholder clarity, the logistic model is the preferred choice. The random forest model may still be useful as a benchmark or ensemble component, but logistic offers the cleanest balance of performance and interpretability.

# High-Level Takeaways from Model Evaluation

# High-Level Takeaways

Summary of observations from performance metrics.

| Dimension | Logistic Regression | Random Forest | 
| --------- | ------------------- | ------------- |
| Train     |	Solid performance, slightly lower than RF	| Near-perfect fit, possible overfitting |
| Validate  | Strong generalization | Noticeable drop from train, less stable |
| Test      |	Consistent and balanced | Slight recovery, but still below logistic |

# Misclassification Analysis

To assess whether the models were simply reproducing the engineered adequacy score, perform a misclassification analysis across adequacy tiers using the unseen test data. 

Tag mis-classified plans with flags for false positives and false negatives using the test set.

```{r}
plans_test_log <- test_results_logistic$predictions

plans_test_rf <- test_results_rf$predictions

# Combine predictions into one data frame.

class_check <- plans_test %>%
  select(ACK_ID, ADEQUACY_IND, ADEQUACY_SCORE) %>%
  left_join(plans_test_log %>%
              select(ACK_ID, predicted_class) %>%
              rename(LOGISTIC_PRED_NUM = predicted_class),
            by = "ACK_ID") %>%
  left_join(plans_test_rf %>%
              select(ACK_ID, predicted_class) %>%
              rename(RF_PRED_NUM = predicted_class),
            by = "ACK_ID") %>%
  mutate(ADEQUACY_IND_NUM = as.numeric(as.character(ADEQUACY_IND)),
         LOGISTIC_FP = LOGISTIC_PRED_NUM == 1 & ADEQUACY_IND_NUM == 0,
         LOGISTIC_FN = LOGISTIC_PRED_NUM == 0 & ADEQUACY_IND_NUM == 1,
         RF_FP = RF_PRED_NUM == 1 & ADEQUACY_IND_NUM == 0,
         RF_FN = RF_PRED_NUM == 0 & ADEQUACY_IND_NUM == 1)
```

Summarize mis-classifications by adequacy score to show whether mis-classified plans are near the threshold (e.g., score of 3 or 4).

```{r}
class_check_summary <- class_check %>%
  group_by(ADEQUACY_SCORE) %>%
  summarise(
    Logistic_FN = sum(LOGISTIC_FN),
    Logistic_FP = sum(LOGISTIC_FP),
    RF_FN = sum(RF_FN),
    RF_FP = sum(RF_FP),
    Total = n()
  ) %>%
  arrange(ADEQUACY_SCORE)

kable(class_check_summary,
      col.names = c("Adequacy Score", "Logistic_FN", "Logistic_FP", "RF_FN", "RF_FP", "Total"),
      caption = "Misclassifications by Model",
      format.args = list(big.mark = ","),
      align = c("l", "r", "r", "r", "r", "r"))
```

Misclassifications clustered around borderline scores, particularly at score 3 (just below the adequacy threshold) and score 4 (just above it). 

Both models flagged several score 3 plans as adequate and several score 4 plans as inadequate, suggesting they were responding to signals in tiered features rather than rigidly following the score cutoff. 

At the extremes (scores 0, 1, and 6), both models classified plans cleanly, reinforcing that they respected strong adequacy signals. 

These patterns validate that the models were not memorizing thresholds but learning nuanced structural patterns embedded in the feature space.

Create faceted bar plot.

```{r}
misclass_long <- class_check_summary %>%
  tidyr::pivot_longer(cols = c(Logistic_FN, Logistic_FP, RF_FN, RF_FP),
               names_to = "Model_Error",
               values_to = "Count") %>%
  mutate(Model = case_when(grepl("^Logistic", Model_Error) ~ "Logistic",
                           grepl("^RF", Model_Error) ~ "Random Forest"),
         Error_Type = case_when(grepl("FN$", Model_Error) ~ "False Negative",
                                grepl("FP$", Model_Error) ~ "False Positive"))

ggplot(misclass_long, aes(x = factor(ADEQUACY_SCORE), y = Count, fill = Error_Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  # geom_vline(xintercept = 4.5, linetype = "dashed", color = "black", linewidth = 0.8) +
  facet_wrap(~ Model) +
  scale_fill_manual(values = c("steelblue", "lightsteelblue")) +
  labs(title = "Misclassifications by Adequacy Score",
       x = "Adequacy Score",
       y = "Count",
       fill = "Error Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
        plot.title = element_text(size = 14, face = "bold"))
```

# Conclusion

The models demonstrate genuine learning from the feature space, not rote reproduction of the adequacy score.

Their misclassification patterns reflect sensitivity to structural nuance, especially near the adequacy boundary, and reinforce their diagnostic value in modeling. This validates their use for stakeholder-facing classification.

Save key data frames for final report.

```{r}
saveRDS(df_all_metrics, "../data/model_metrics.rds")

saveRDS(class_check_summary, "../data/check_class_summary.rds")
```
